Natural sources of vagueness and their implications

### 2017-05-08 book_chapter/paper.tex
Fine-tuned reference games section. Finished conclusion.

% \item We do not play only one language game in our existence; there is a plurality of them and which one an agent is engaged in at a particular moment is never clearly identified, neither are the exact benefits one might take out of it by choosing a certain behavior over another;
% \item These are furthermore not fixed in time; old language games fall out of fashion or stop being useful, and new ones emerge all the time.

Linguistic interaction, as captured by a sim-max game, has a rather narrow scope. There is a fixed prior, a fixed utility function and a fixed similarity measure between states. 
If the sender observes a state $t$ and the receiver guesses state $t'$ the utility of that interaction is proportional to the similarity between $t$ and $t$. 
Here is a natural example of a situation captured by a sim-max model: Kiki wants Bubu to draw her dissertation cover in a particular shade of blue $t$; the closer Bubu's color choice matches Kiki's idea, the better. Here Kiki wants to \emph{describe} the color for Bubu to pick out or realize from the whole set of possibilities. 
It's a natural thing to want to do with language. 
But a lot of communication with vague and contiguously variable properties is rather different. 
In the general spirit of Wittgenstein's method of arranging multiple language games side by side for a more encompassing picture, we should look at some more cases of language use as well. 
Let us therefore consider a continuum of cases of referential communication.

A \emph{reference game} is a model of a situation in which the speaker has a particular object in mind and uses a description in order for the receiver to pick this object and no other object. 
One crucial difference to descriptive communication and sim-max games is that close will not be close enough: if Kiki needs a long nail, a nail a little longer or shorter may be good enough; but if Bubu needs the long, slightly bend nail with the flat top (which she previously disassembled from her art installation), no other nail will do.

@@

The picture sketched in the last paragraph neglects learning from observation and generalization across contexts. 
For rationality, we need to consider a rational agent's ability to learn from previous experience and to generalize from  the past to the future. 
For evolutionary adaptation, the story so far misses the key fact that languages and practices have to be learnable in appropriate time from the actual observables, lest what is learned and further transmitted will simply be a different language or practice \parencite[\emph{e.g.}][]{KirbyGriffith2014:Iterated-Learni}.
But even if every language learner in history had been perfectly rational, it is extremely unlikely that anyone ever succeeded in reducing all uncertainty about language use and say with confidence: this is the threshold that truly separates heaps from non-heaps no matter what. 
For one, there may simply not be enough time in the lifespan of a rational learner to observe enough instances of language use for certainty about crisp denotations. 
For another, what needs to be learned likely changes over time. 
The optimal use for gradable properties hinges crucially on the prior over states \parencite[\emph{e.g.}][]{Franke2012:Scales-Salience,Franke2012:On-Scales-Salie,QingFranke2014:Gradable-Adject,LassiterGoodman2015:Adjectival-vagu}. 
If the prior captures real world attributes, such as the probability of an individual's body height, these attributes will not only be hard to know for certain, but they will also vary over time. 
They may also depend on location: do we use our language exclusively in this village or do we travel beyond? 
If so, the priors that determine the goodness of a convention for one agent need not be the same as those for another. 
Going beyond the priors for property degrees, the probability of seeing a particular language game, e.g.~a particular context $c$ in some kind of reference game, is also crucial for determining globally efficient linguistic behavior. 
But here it is even more inconceivable how a rational learner could acquire a fairly confident estimate over how often a particular language game will be played. 
The list of sources of natural and seemingly insurmountable uncertainty about the parameters necessary to determine locally or globally optimal strategies is much longer than hinted at here. 
Still, it should be clear enough that a finite existence, a sparse learning input and the inherent fuzziness of the linguistic context itself (its temporal, local and social variability) is not a faul of rationality or the mechanism of selection of successful linguistic practices.

What about meaning and vagueness in a universe of multiple language games and heavy uncertainty? 
We could look for meaning at several levels. 
Firstly, as before, there is the actual behavior of a single agent in each actualized language game. 
As mentioned above, behavior that strictly maximizes expected utility under uncertainty may be resource heavy, so it might be compatible with local strategic rationality that agents' production choices are stochastic (with the caveat that in the real world we would never know because we cannot ever elicit a production in the exact same circumstances in a constantly changing environment including the mind of a speaker).
Secondly, if we look at behavior across many game types and contexts, there is also the level of an agent's internal theory of how words and phrases are likely to be used, conditional on a given context. 
Notice that a single agent's rational beliefs about linguistic practices or linguistics meaning may well have to reflect the actual stochasticity: under natural assumptions about information loss, the best belief for prediction matches the actual distribution in the real world \parencite[\emph{e.g.}][]{VehtariOjanen2012:A-survey-of-Bay}.
Either one of these two levels, the agent's behavior or linguistic generalizations, could be the foundation for a notion of emerging abstract meaning as constituted by the practices of a population at a particular time.
If both agents' rational behavior and rational beliefs are plausibly fuzzy, why should a concept of supervening meaning be crisp? 
Maybe there is a reason.
The main question to be answered is even more general: given a universe of (rational) language users who learn from finite experience in a changing and stochastically complex arrangement of multiple language games, is a notion of linguistic meaning a theoretically useful and empirically explanatory concept in this picture in such a way that if that notion entailed vague meanings the whole view would have disastrously little explanatory bite?

Taken together, when we look at the picture of multiple language games side by side, the question of what rational or optimal behavior is in any one of them, or the question of why human language users do not more closely approximate rational or optimal local behavior more closely seem entirely misplaced.
What seems more pressing is to answer how the interaction of individual learning, individual local optimization and general pressure for successful communication at a cross-situational level lead to the shaping of linguistic practices in an open-ended landscape of language games.
How to attain locally sub-optimal play in a single language game is not a theoretical puzzle we should take serious, if what really matters is a global multi-game perspective.
Adopting this perspective does not \emph{solve} the problem of how vagueness is compatible with rationality. 
But it suggest that it might be an inflated issue that we need not or should not worry about, because it will not matter for a more general and pressing issue in a more realistic and encompassing picture of language, its use and evolution. The following section will expand on this.

@@

We tried to show that, if we pragmatically acknowledge the complex and dynamic environment that surrounds language learning and use, we find many plausible sources of uncertainty that can lead to vague signal use.
Furthermore, if we broaden our notion of rationality beyond mere maximization of expected utility, we also find reasons to welcome the positive aspects of these processes and to see their acceptance as a perfectly rational choice.
Irrational would be to ignore the finite 
The investigation doesn't end here, though.
The research discussed here suggests that although vagueness can be beneficial, that is the case only in moderation: too much can lead to babbling equilibria where no information is transmitted.
Thus, for agents capable of adjusting the level of vagueness in their strategies, questions remain regarding how this can be modeled and what factors influence the amount of vagueness that is optimal for a given situation.
Other challenges include how to incorporate the intuitions about the relevance of the aspects enumerated above into our models.

>>>>

The type of linguistic interaction captured by a sim-max game has its own particular scope.
There is a fixed prior, a fixed utility function and a fixed similarity measure between states. 
If the sender observes a state $t$ and the receiver guesses state $t'$ the utility of that interaction is proportional to the similarity between $t$ and $t'$.
Here is a natural example of a situation captured by a sim-max model: Kiki wants Bubu to draw her dissertation cover in a particular shade of blue; the closer Bubu's color choice matches Kiki's idea, the better. Here Kiki wants to \emph{describe} the color for Bubu to pick out or realize from the whole set of possibilities. 
It's a natural thing to want to do with language. 
But other types of communication with vague and contiguously variable properties are possible. 
%In the general spirit of Wittgenstein's method of arranging multiple language games side by side for a more encompassing picture, we should look at some more cases of language use as well. 
%Let us therefore consider a continuum of cases of referential communication.
A \emph{reference game} is a model of a situation in which the speaker has a particular object in mind and uses a description in order for the receiver to pick this object and no other object. 
One crucial difference to descriptive communication and sim-max games is that close will not be close enough: if Kiki needs a long nail, a nail a little longer or shorter may be good enough; but if Bubu needs the long, slightly bend nail with the flat top (which she previously disassembled from her art installation), no other nail will do.

@@

The picture just sketched neglects learning from observation, as well as generalization across contexts. 
For rationality\todo{clarify}, we need to consider a rational agent's ability to learn from previous experience and to generalize from  the past to the future. 
For evolutionary adaptation, the story so far misses the key fact that languages and practices have to be learnable in appropriate time from the actual observables, lest what is learned and further transmitted will simply be a different language or practice \parencite[\emph{e.g.}][]{KirbyGriffith2014:Iterated-Learni}.
But even if every language learner in history had been perfectly rational, it is extremely unlikely, as was discussed in the previous section, that anyone ever succeeded in reducing all uncertainty about language use and say with confidence: this is the threshold that truly separates heaps from non-heaps no matter what. 
For one, there may simply not be enough time in the lifespan of a rational learner to observe enough instances of language use for certainty about crisp denotations. 
For another, what needs to be learned likely changes over time. 
The optimal use for gradable properties hinges crucially on the prior over states \parencite[\emph{e.g.}][]{Franke2012:Scales-Salience,Franke2012:On-Scales-Salie,QingFranke2014:Gradable-Adject,LassiterGoodman2015:Adjectival-vagu}. 
If the prior captures real world attributes, such as the probability of an individual's body height, these attributes will not only be hard to know for certain, but they will also vary over time. 
They may also depend on location: do we use our language exclusively in this village or do we travel beyond? 
If so, the priors that determine the goodness of a convention for one agent need not be the same as those for another. 
Going beyond the priors for property degrees, the probability of seeing a particular context $c$ in some kind of reference game, is also crucial for determining globally efficient linguistic behavior. 
But here it is even more inconceivable how a rational learner could acquire a fairly confident estimate over how often the upcoming reference games will have a particular context or another. 
The list of sources of natural and seemingly insurmountable uncertainty about the parameters necessary to determine locally or globally optimal strategies is much longer than hinted at here. 
Still, it should be clear enough that a finite existence, a sparse learning input and the inherent fuzziness of the linguistic context itself (its temporal, local and social variability) is not a fault of rationality on the part of the agent, or the mechanism of selection of successful linguistic practices.

What about meaning and vagueness in a universe of multiple language games and heavy uncertainty?
We do not play only one language game in our existence; there is a plurality of them and which one an agent is engaged in at a particular moment is never clearly identified, neither are the exact benefits one might take out of it by choosing a certain behavior over another.
These are furthermore not fixed in time; old language games fall out of fashion or stop being useful, and new ones emerge all the time. 
We could look for meaning at several levels. 
Firstly, as before, there is the actual behavior of a single agent in each actualized language game. 
As mentioned above\todo{where?}, behavior that strictly maximizes expected utility under uncertainty may be resource heavy, so it might be compatible with local strategic rationality that agents' production choices are stochastic (with the caveat that in the real world we would never know because we cannot ever elicit a production in the exact same circumstances in a constantly changing environment including the mind of a speaker).
Secondly, if we look at behavior across many game types and contexts, there is also the level of an agent's internal theory of how words and phrases are likely to be used, conditional on a given context. 
Notice that a single agent's rational beliefs about linguistic practices or linguistic meaning may well have to reflect the actual stochasticity: under natural assumptions about information loss, the best belief for prediction matches the actual distribution in the real world \parencite[\emph{e.g.}][]{VehtariOjanen2012:A-survey-of-Bay}.
Either one of these two levels, the agent's behavior or linguistic generalizations, could be the foundation for a notion of emerging abstract meaning as constituted by the practices of a population at a particular time.
If both agents' rational behavior and rational beliefs are plausibly fuzzy, why should a concept of supervening meaning be crisp? 
Maybe there is a reason.
The main question\todo{needs more argumentation, if we even want to go there} to be answered is even more general: given a universe of (rational) language users who learn from finite experience in a changing and stochastically complex arrangement of multiple language games, is a notion of linguistic meaning a theoretically useful and empirically explanatory concept in this picture in such a way that if that notion entailed vague meanings the whole view would have disastrously little explanatory bite?

Taken together, when we look at the picture of multiple language games side by side, the question of what rational or optimal behavior is in any one of them, or the question of why human language users do not more closely approximate rational or optimal local behavior more closely seem entirely misplaced.
What seems more pressing is to answer how the interaction of individual learning, individual local optimization and general pressure for successful communication at a cross-situational level lead to the shaping of linguistic practices in an open-ended landscape of language games.
How to attain locally sub-optimal play in a single language game is not a theoretical puzzle we should take serious, if what really matters is a global multi-game perspective.
Adopting this perspective does not \emph{solve} the problem of how vagueness is compatible with rationality. 
But it suggest that it might be an inflated issue that we need not or should not worry about, because it will not matter for a more general and pressing issue in a more realistic and encompassing picture of language, its use and evolution.

@@

We tried to show that, if we pragmatically acknowledge the complex and dynamic environment that surrounds language learning and use, we find many plausible sources of uncertainty that can lead to vague signal use.
We supported this claim by giving an overview of recent research that tries to accommodate for vagueness in the context of sim-max games, and by discussing, in more general terms, how these considerations are relevant for reference games, and in the broader context of language as a multiplicity of language-games.
Furthermore, we argued that if we broaden our notion of rationality beyond mere maximization of expected utility, we also find reasons to welcome the positive aspects of these processes and to see their acceptance as a perfectly rational choice.
Irrational would be to ignore the finite nature of our existence and the constraints that condition language learning and use.
Our challenge as researchers is of course not only to draw attention to the possibilities, but also to characterize which particular constraints give rise to vague signal use and how.

The investigation doesn't end here, though.
The research discussed here suggests that although vagueness can be beneficial, that is the case only in moderation: too much can lead to babbling equilibria where no information is transmitted.
Thus, for agents capable of adjusting the level of vagueness in their strategies, questions remain regarding how this can be modeled and what factors influence the amount of vagueness that is optimal for a given situation.
Other challenges include how to incorporate into our models the intuitions about the importance of speed of convergence, adaptability, and the promotion of group homogeneity.
Finally, perhaps the biggest of them all is how to be able to focus on one language-game without losing sight of the ecosystem of games that need to be taken into account as well.
Whatever solutions we find to these challenges, we believe that a notion of rationality cannot be strongly tied to mere utility maximization in a particular game, but has to consider an agents' need to optimize his behavior to navigate the plurality of uncertainties that besiege him.
